---
title: "Mutations and Lineage trees"
author: "Milca Tarshish"
date: "12/3/2019"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#### Loading libraries:
```{r, message=F}
library(alakazam)
library(igraph)
library(shazam)
library(crayon)
```

#### Processing data:  
Loading data, and subseting one clone out of it:  
```{r}
data("ExampleDb")
#clone_id<-"3128"  # This clone has 100 sequences
clone_id<-"3110"  # This clone has 26 sequences
clone_db<-subset(ExampleDb,CLONE==clone_id)
dim(clone_db)
```
Generating a ChangeoClone object for lineage construction:  
```{r}
clone <- makeChangeoClone(clone_db, text_fields=c("SAMPLE", "ISOTYPE"), num_fields="DUPCOUNT")
dim(clone@data)
```

Building a Phylip lineage tree out of the ChangeoClone object above:  
```{r}
dnapars_exec <- "c:\\Users\\milcat\\phylip-3.698\\exe\\dnapars.exe"
graph <- buildPhylipLineage(clone, dnapars_exec, rm_temp=TRUE)
```
Giving sequence numbers to the graph vertexes, and building a lookup table of sequence_id versus sequence_number:  
(Sequence number are much easier for me to track than a long sequence id).   
```{r}
vertex_lookup<-data.frame(seq_num=c(1:(vcount(graph)-1)),
                          seq_id=as_data_frame(graph,what="edges")$to)
                     
vertex_lookup  
```
Building a data frame called "graph_df" from the graph and lookup table above:
```{r}
graph_df<-data.frame(sequence_num=c(1:(vcount(graph)-1)),
                     parent_sequence_num=rep("NA",vcount(graph)-1),  # placeholder for this column, putting NA at this stage.
                     sequence_id=as_data_frame(graph,what="edges")$to, 
                     parent_sequence_id=as_data_frame(graph,what="edges")$from
                     )
graph_df$parent_sequence_num<-vertex_lookup[match(graph_df$parent_sequence_id,vertex_lookup$seq_id),"seq_num"]
graph_df
```
Generating data frame with tableEdges function (found this function later on, so just using it to compare my results):  
```{r}
tableEdges(graph, "name")
```

Merging the 2 data frames:  
1. graph_df  
2. data frame of the clone object  
```{r}
merge_df<-merge(graph_df,clone@data,by.x="sequence_id", by.y="SEQUENCE_ID",all=TRUE)
```
Adding to the merged data frame - the parrent sequence for each sequence
```{r}
merge_df$PARENT_SEQUENCE<-merge_df[match(merge_df$parent_sequence_num,merge_df$sequence_num),"SEQUENCE"]
#replacing missing sequence (where value is na) with germline sequence
merge_df[c(which(is.na(merge_df[,"SEQUENCE"]))),"SEQUENCE"]<-clone@germline
#replace mising parent sequences (where value is na) with germline sequence
merge_df[c(which(is.na(merge_df[,"PARENT_SEQUENCE"]))),"PARENT_SEQUENCE"]<-clone@germline
```
plotting the graph:  
```{r}
V(graph)$color <- "steelblue"
V(graph)$color[V(graph)$name == "Germline"] <- "lightgreen"
V(graph)$color[grepl("Inferred", V(graph)$name)] <- "purple"
V(graph)$seq_num<-vertex_lookup[match(V(graph)$label,vertex_lookup$seq_id),"seq_num"]
V(graph)$seq_num[V(graph)$name == "Germline"]<-"Germline"
E(graph)$label <- ""
par(mar=c(0, 0, 0, 0) + 0.8)
plot(graph, layout=layout_as_tree, edge.arrow.mode=0, vertex.frame.color="black",
     vertex.label.color="black", vertex.size=20, vertex.label=V(graph)$seq_num,
     main=paste("Lineage tree for clone clone_id",clone_id))
legend("topleft", c("Germline", "Inferred", "Sample"), 
       fill=c("lightgreen", "purple", "steelblue"), cex=0.75)
```

Adding to merge_df - the expected and obserevd mutations:  
```{r}
obs_df<-observedMutations(merge_df, sequenceColumn="SEQUENCE",
                            germlineColumn="PARENT_SEQUENCE",
                            frequency=TRUE,
                            nproc=1)
exp_df<-expectedMutations(merge_df, sequenceColumn="SEQUENCE",
                            germlineColumn="PARENT_SEQUENCE",
                            nproc=1)
obs_df[,c("sequence_num","parent_sequence_num","MU_FREQ_SEQ_R", "MU_FREQ_SEQ_S")]
exp_df[,c("sequence_num","parent_sequence_num","MU_EXPECTED_SEQ_R", "MU_EXPECTED_SEQ_S")]
```


